Automatic Detection of Bad Programming Habits in Scratch: A Preliminary Study

### 2014-04-23 paperFIE_en.tex
commenting the results. balancing the columns in the final page. passing aspell. It is now a release candidate

%\documentclass[a4paper,10pt]{article}
\documentclass[conference]{format/IEEEtran}
%\usepackage[utf8x]{inputenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{url}

@@

However, there are reports that indicate that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations. In our work as instructors with high school students, we have detected these and other bad practices, such as the repetition of code and object naming, on a regular basis. This paper focuses on verifying whether these issues can generally be found in the Scratch community, by analyzing a large number of projects available on the Scractch community website.

@@

In order to test the functionality of the developed plug-ins and to verify whether these behaviors are common in projects found in the Scratch repository, we proceeded to randomly download 100 projects from the repository. As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created for this version. A replication package of the projects under study, and the specific version of the plug-ins used in it can be obtained publicly\footnote{FIXME: URL \url{http://gsyc.urjc.es/~grex/repro/2014-fie-scratch}}.

@@

The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects, as shown in table \ref{table:results}.

Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, 79\% of the analyzed projects contained at least one non-renamed \textit{SpriteX} character.

As for the repetition code, 62\% of the analyzed projects contained at least one repeated program. In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project.

FIXME: here comes the table with some more results. Important: to note how the distribution (talk about mean and median).

\begin{table}
\begin{center}
  \begin{tabular}{ | p{2.5cm} | p{1cm} | p{1.2cm} | p{1cm} |}
   \hline
     & Default names & Duplicated scripts & Defined blocks \\ \hline\hline
    Projects incurring in & 79 & 62 & 17 \\ \hline
    Mean & 5.94 & 7.23 & 1.11 \\ \hline
    Median & 3 & 2 & 0 \\ \hline
    Maximum & 67 & 71 & 25 \\
    \hline    
  \end{tabular}
\end{center}
\caption{Table of results}
\label{table:results}
\end{table}

>>>>

%\documentclass[a4paper,10pt]{article}
\documentclass[conference]{format/IEEEtran}
%\usepackage[utf8x]{inputenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{url}
\clubpenalty = 10000
\widowpenalty = 10000
\displaywidowpenalty = 10000
\usepackage{flushend}

@@

However, there are reports that indicate that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations. In our work as instructors with high school students, we have detected these and other bad practices, such as the repetition of code and object naming, on a regular basis. This paper focuses on verifying whether these issues can generally be found in the Scratch community, by analyzing a large number of projects available on the Scratch community website.

@@

In order to test the functionality of the developed plug-ins and to verify whether these behaviors are common in projects found in the Scratch repository, we proceeded to randomly download 100 projects from the repository. As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created for this version. A replication package of the projects under study, and the specific version of the plug-ins used in it can be obtained publicly\footnote{\url{http://gsyc.urjc.es/~grex/repro/2014-fie-scratch}}.

@@

The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects, as shown in Table~\ref{table:results}.

\begin{table}
\begin{center}
  \begin{tabular}{ | l | r | r | r |}
   \hline
              & Default names & Duplicated scripts & Blocks defined \\ \hline\hline
    Projects & 79 & 62 & 17 \\ \hline
    Mean & 5.94 & 7.23 & 1.11 \\ \hline
    Median & 3 & 2 & 0 \\ \hline
    Maximum & 67 & 71 & 25 \\
    \hline    
  \end{tabular}
\end{center}
\caption{Table of results}
\label{table:results}
\end{table}

Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, 79\% of the analyzed projects contained at least one non-renamed \textit{SpriteX} character. The maximum number of non-renamed \textit{SpriteX}s found has been 67, which makes the code very hard to maintain (i.e., one could consider this almost as \emph{code obfuscation}). In mean, almost 6 \textit{SpriteX}s can be found in a project, while the median is 3. Our results show that this is a severe problem in the analyzed projects in Scratch, as in addition to the fact that re-naming occurs in very few projects, the number of elements used in the others are significant in number; being those programs are very hard to understand, its maintenance and reuse are hindered.

As for the repetition code, 62\% of the analyzed projects contained at least one repeated program. In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project. The mean of duplicated scripts and blocks defined provides us insight about how difficult is for the authors of the analyzed projects to use abstraction and modularization: duplicated code appears over six times more than blocks.
